---
title: "exercice"
format: html
---

# Task 1

## Import data
```{r}
library(sf)
library("readr")
library(terra)

wildschwein <- read_delim("wildschwein_BE_2056.csv", ",") |>
    st_as_sf(coords = c("E", "N"), crs = 2056, remove = FALSE)
Feld <- read_sf("Feldaufnahmen_Fanel.gpkg")
my_tif <- rast("vegetationshoehe_LFI.tif")
plot(my_tif)
```
## Questions about dataset
a) What information does the dataset contain?
Koordinaten von verschiedenen Ackerkulturen
b) What is the geometry type of the dataset (possible types are: Point, Lines and Polygons)?
Polygon
c) What are the data types of the other columns?
Double and character 
d) What is the coordinate system of the dataset?
CH1903+ / LV95

# Task 2
```{r}
#filter wild boar data
library(lubridate)
library(dplyr)
wildschwein <- wildschwein %>%
  filter(month(DatetimeUTC) %in% c(5, 6))
#st join 
joineddata<- st_join(wildschwein, Feld)
```

# Task 3 
```{r}
#visualisation
#which fields was more visited by the wildschwein 
library(ggplot2)

joineddata %>%
  st_drop_geometry() %>%
  count(TierName, Frucht, sort = TRUE) %>%
  ggplot(aes(x = Frucht, y = n, fill = TierName)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    x = "Polygon / crop type",
    y = "Number of GPS points",
    fill = "Wild boar"
  )
#temporal analysis 
library(dplyr)
library(lubridate)
library(ggplot2)
library(sf)

wildschwein_clean <- joineddata %>%
  st_drop_geometry() %>%
  mutate(
    date = as.Date(DatetimeUTC),
    Frucht = ifelse(is.na(Frucht), "Unknown", as.character(Frucht))
  )

# Keep only the most common crop types, group the rest as "Other"
top_frucht <- wildschwein_clean %>%
  count(Frucht, sort = TRUE) %>%
  slice_head(n = 6) %>%
  pull(Frucht)

wildschwein_plot <- wildschwein_clean %>%
  mutate(
    Frucht_group = ifelse(Frucht %in% top_frucht, Frucht, "Other")
  )

# Dominant crop type per day and wild boar
daily_dominant <- wildschwein_plot %>%
  count(TierName, date, Frucht_group) %>%
  group_by(TierName, date) %>%
  slice_max(n, n = 1, with_ties = FALSE) %>%
  ungroup()

ggplot(daily_dominant, aes(x = date, y = TierName, fill = Frucht_group)) +
  geom_tile(height = 0.75) +
  labs(
    x = "Date",
    y = "Wild boar",
    fill = "Dominant polygon / crop type",
    title = "Dominant visited crop type per day and wild boar"
  ) +
  theme_minimal()
```

# Task 4 
already done in 1 

# Task 5 
```{r}
library(terra)
# 1. Convertir les points sf en SpatVector
wildschwein_vect <- terra::vect(wildschwein)

# 2. Extraire la valeur du raster pour chaque point
tif_values <- terra::extract(my_tif, wildschwein_vect)

# 3. Enlever la colonne ID créée par terra
tif_values <- tif_values %>%
  select(-ID)

# 4. Ajouter la valeur extraite au dataset original
new_join <- bind_cols(wildschwein, tif_values)


```
visualisation
```{r}
library(sf)
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidyr)
library(scales)

# Préparer les données
plot_data <- new_join %>%
  st_drop_geometry() %>%
  mutate(
    DatetimeUTC = as.POSIXct(DatetimeUTC, tz = "UTC"),
    hour_round = hour(round_date(DatetimeUTC, unit = "hour")),
    
    vegetation_class = cut(
      vegetationshoehe_LFI,
      breaks = c(-Inf, 5, 10, 15, 20, 25, Inf),
      labels = c(
        "0-5 m",
        "5-10 m",
        "10-15 m",
        "15-20 m",
        "20-25 m",
        ">25 m"
      ),
      right = FALSE
    ),
    
    vegetation_class = as.character(vegetation_class),
    vegetation_class = ifelse(is.na(vegetation_class), "No data", vegetation_class)
  )

# Compter les points GPS par sanglier, heure et classe de végétation
hourly_counts <- plot_data %>%
  count(TierName, hour_round, vegetation_class)

# Compléter les heures manquantes
hourly_counts <- hourly_counts %>%
  complete(
    TierName,
    hour_round = 0:23,
    vegetation_class,
    fill = list(n = 0)
  )

# Calculer les pourcentages
hourly_props <- hourly_counts %>%
  group_by(TierName, hour_round) %>%
  mutate(
    percentage = n / sum(n)
  ) %>%
  ungroup()

# Ordre des classes dans la légende
hourly_props <- hourly_props %>%
  mutate(
    vegetation_class = factor(
      vegetation_class,
      levels = c("0-5 m", "5-10 m", "10-15 m", "15-20 m", "20-25 m", ">25 m", "No data")
    )
  )

# Graphique
percent<- ggplot(hourly_props, aes(x = hour_round, y = percentage, fill = vegetation_class)) +
  geom_col(width = 1) +
  facet_wrap(~ TierName, nrow = 1) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_x_continuous(breaks = seq(0, 23, by = 5)) +
  labs(
    x = "Time (rounded to the nearest hour)",
    y = "Percentage",
    fill = "Vegetation height",
    title = "Vegetation height use by wild boar over the day"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 14, face = "bold"),
    legend.position = "bottom"
  )
percent
```

